---
title: "MOTA_PCA_GUT_code"
author: "Pierre FOUCAULT"
date: "12/8/2021"
output: html_document
---

Welcome to the Multivariate Omics Trajectory Analysis (MOTA) code.

## PACKAGES
##|

To begin, let's load all the necessary packages 

```{r}
library(phyloseq)
library(qiime2R)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(dplyr)
library(mixOmics)
library(readr)
library(stringr)

#If not already downloaded
#BiocManager::install("_____")
```

## LOAD DATA
##|

Then let's load the data associated with the manuscript.
Two datasets are expected, two  mixOmics-friendly dataset, one for the micorbiome (metabarcoding, 16S rRNA) data and the other for the metabolome (mass spectrometry, LC-MS) data.

```{r}
#### Microbiome data ####

Gadn<-read_delim("gut_adn_mixomics_2.csv",";", escape_double = FALSE, trim_ws = TRUE) %>%
  remove_rownames %>% column_to_rownames(var="Names") #View(Gadn)

MetadataGUT_log<-read_delim("MetadataGUT_D.csv",";", escape_double = FALSE, trim_ws = TRUE) %>%
  remove_rownames %>% column_to_rownames(var="Names") #View(MetadataGUT_log)

#### Metabolome data ####

G_log<-read_delim("essai_gut_metabo_norm_log.csv",";", escape_double = FALSE, trim_ws = TRUE) %>% remove_rownames %>% column_to_rownames(var="Names") #View(G_log)

MetadataGUT_log<-read_delim("MetadataGUT_D.csv",";", escape_double = FALSE, trim_ws = TRUE) %>%
  remove_rownames %>% column_to_rownames(var="Names") #View(MetadataGUT_log)
```

## METRICS
##|

This pipeline is based on ordinate object for the microbiome data. This script is dedicated to the PCA ordination method.

Ordinated objects can contain many axis, up to a point where a non-informative % 
of the remaining ‰ of Explained variance is carried. 
The "Threshold" variable set the ‰ of Expl. Var you want to work with.

The function used to compute the average distance between two time points can be chosen
with the "distance_metrics" variable.

```{r}
Threshold= 0.9
distance_metrics="median"
```

```{r}
if (distance_metrics=="median") {
  distance_metrics=median
} else distance_metrics=mean
```

## MICROBIOME

### PCA

let's begin with the microbiome data. The goal is to obtained a dataframe containing cumulative distances
for each experimental condition (0 and 100 µg Cha/L here).

To obtain theses distances, an ordination must be performed.
The ‰ of Expl. Var. and the coordinates of each axis will be stored.

```{r}
PCAGut_adnmedaka.pca<-pca(Gadn)
View(PCAGut_adnmedaka.pca$cum.var)
```

### AXIS NEEDED

The following loop process the number of axis needed to achieved the designated ‰ of Expl. Var. .

```{r}
ExplVarTot_pcamicrobiome=as.data.frame(PCAGut_adnmedaka.pca$cum.var)[nrow(as.data.frame(PCAGut_adnmedaka.pca$cum.var)),1] #ExplVarTot_pcamicrobiome
as.data.frame(PCAGut_adnmedaka.pca$cum.var)

if (ExplVarTot_pcamicrobiome < Threshold) {
  ExplVarTot_pcamicrobiome=0
  nb_axis_pcamicrobiome=1
  while (ExplVarTot_pcamicrobiome<Threshold) {
    nb_axis_pcamicrobiome=nb_axis_pcamicrobiome+1
    PCAGut_adnmedaka.pca<-pca(Gadn,ncomp=nb_axis_pcamicrobiome)
    ExplVarTot_pcamicrobiome=as.data.frame(PCAGut_adnmedaka.pca$cum.var)[nrow(as.data.frame(PCAGut_adnmedaka.pca$cum.var)),1]
    Axis_needed_pcamicrobiome<-nb_axis_pcamicrobiome
    Axis_needed_end_pcamicrobiome<-nb_axis_pcamicrobiome+2
  }
  print(paste(nb_axis_pcamicrobiome,"Axis are needed to explain",ExplVarTot_pcamicrobiome,"of the total variance"))
  ExplVarTot_pcamicrobiome=0
  nb_axis_pcamicrobiome=0
}
```

### CENTROIDS CORDINATES

In order to compute the distance between each time point, a unique set of coordinates must me associated with each group (Time_treatment here). Centroids can be processed to work out these coordinate. 
First, we need to create a new data frame containing all coordinates and the metadata needed.
A reduced dataset can be cretaed by retaining only the necessary axis fixed earlier.

```{r}
newdf_pcamicrobiome<-data.frame(MetadataGUT_log$Cond2,MetadataGUT_log$Treatment,PCAGut_adnmedaka.pca$variates$X) #View(newdf_pcamicrobiome)
newdf_pcamicrobiome_split<-split(newdf_pcamicrobiome,f = MetadataGUT_log$Treatment) 
#View(newdf_pcamicrobiome_split$`0`) 
#View(newdf_pcamicrobiome_split$`100`)
```

At this step, the dataset need to be split to compute all the distances for each treatment.
The variable "Exposure" form the metadata will be used.

```{r}
newdf_pcamicrobiome_0<-newdf_pcamicrobiome_split$`0` %>% group_by(newdf_pcamicrobiome_split$`0`$MetadataGUT_log.Cond2) %>%
  dplyr::summarise(across(starts_with("PC"),.fns=distance_metrics)) #View(newdf_pcamicrobiome_0)

newdf_pcamicrobiome_100<-newdf_pcamicrobiome_split$`100` %>% group_by(newdf_pcamicrobiome_split$`100`$MetadataGUT_log.Cond2) %>%
  dplyr::summarise(across(starts_with("PC"), .fns=distance_metrics)) #View(newdf_pcamicrobiome_100)
```

The first time point (d0_0) is only associated with the first condition while
being the starting point of all treatments in the same time.
It needs to be added to each treatment.

```{r}
#ajouter le d0_0 au dataframe exposé
names(newdf_pcamicrobiome_0)[1] <-"Time_Exposure"
names(newdf_pcamicrobiome_100)[1] <-"Time_Exposure"
newdf_pcamicrobiome_100<-rbind(newdf_pcamicrobiome_0[1,],newdf_pcamicrobiome_100) #View(newdf_pcamicrobiome_100)
```

### EUCLIDEAN NORM

The distances can then be processed by computing the euclidean norm between each group for all treatments.
Let's start with the non-exposed/control samples.

#### df_microbiome_0

Three new columns will be added : 
  - a name/label for all cumputed-distances
  - the euclidean norm
  - the cumulative euclidean norm in %

```{r}
#col distance_label
Nlabel_pcamicrobiome=1
distance_label_pcamicrobiome_0<-as.character(newdf_pcamicrobiome_0[1,1]) #View(distance_label_pcamicrobiome_0)
while (Nlabel_pcamicrobiome<nrow(newdf_pcamicrobiome_0)) {
  distance_label2_pcamicrobiome_0<-as.character(paste(newdf_pcamicrobiome_0[Nlabel_pcamicrobiome,1],"_",newdf_pcamicrobiome_0[Nlabel_pcamicrobiome+1,1]))
  distance_label_pcamicrobiome_0<-c(distance_label_pcamicrobiome_0,distance_label2_pcamicrobiome_0)
  Nlabel_pcamicrobiome=Nlabel_pcamicrobiome+1
}
newdf_pcamicrobiome_0$distance_label<-distance_label_pcamicrobiome_0 #View(newdf_pcamicrobiome_0)

#col euclidean_norm
euclidean_norm_pcamicrobiome_0<-as.numeric(0) #View(euclidean_norm_pcamicrobiome_0)
i=2
while (i<=nrow(newdf_pcamicrobiome_0)) {
  sum_distance=0
  for (n in 2:(Axis_needed_pcamicrobiome+1)) {
    distance_0<-sum((newdf_pcamicrobiome_0[i,n]-newdf_pcamicrobiome_0[i-1,n])^2)
    sum_distance=sum_distance+distance_0
  }
  euclidean_norm_pcamicrobiome_0<-c(euclidean_norm_pcamicrobiome_0,sqrt(sum_distance))
  print(paste("The total distance",i,"is",sqrt(sum_distance),"For",(n-1),"axis"))
  i=i+1
}
print(euclidean_norm_pcamicrobiome_0)
newdf_pcamicrobiome_0$euclidean_norm_pcamicrobiome_0<-euclidean_norm_pcamicrobiome_0 #View(newdf_pcamicrobiome_0)
Axis_needed_pcamicrobiome
#col cumdistance
cumdistance_pcamicrobiome_0<-as.numeric(0)
i=2
while (i<=nrow(newdf_pcamicrobiome_0)) {
  distancep_0<-sum(((newdf_pcamicrobiome_0$euclidean_norm_pcamicrobiome_0[i]/sum(euclidean_norm_pcamicrobiome_0))*100)+cumdistance_pcamicrobiome_0[i-1])
  cumdistance_pcamicrobiome_0<-c(cumdistance_pcamicrobiome_0,distancep_0)
  i=i+1
}
print(cumdistance_pcamicrobiome_0)
newdf_pcamicrobiome_0$cumdistance_pcamicrobiome_0<-cumdistance_pcamicrobiome_0 #View(newdf_pcamicrobiome_0)
```

#### df_microbiome_100

Let's perform the same process  to the exposed-samples.

```{r}

#col distance_label
Nlabel_pcamicrobiome=1
distance_label_pcamicrobiome_100<-as.character(newdf_pcamicrobiome_100[1,1]) #View(distance_label_pcamicrobiome_100)
while (Nlabel_pcamicrobiome<nrow(newdf_pcamicrobiome_100)) {
  distance_label2_pcamicrobiome_100<-as.character(paste(newdf_pcamicrobiome_100[Nlabel_pcamicrobiome,1],"_",newdf_pcamicrobiome_100[Nlabel_pcamicrobiome+1,1]))
  distance_label_pcamicrobiome_100<-c(distance_label_pcamicrobiome_100,distance_label2_pcamicrobiome_100)
  Nlabel_pcamicrobiome=Nlabel_pcamicrobiome+1
}
newdf_pcamicrobiome_100$distance_label<-distance_label_pcamicrobiome_100 #View(newdf_pcamicrobiome_100)

#col euclidean_norm
euclidean_norm_pcamicrobiome_100<-as.numeric(0) #View(euclidean_norm_pcamicrobiome_100)
i=2
while (i<=nrow(newdf_pcamicrobiome_100)) {
  sum_distance=0
  for (n in 2:(Axis_needed_pcamicrobiome+1)) {
    distance_100<-sum((newdf_pcamicrobiome_100[i,n]-newdf_pcamicrobiome_100[i-1,n])^2)
    sum_distance=sum_distance+distance_100
  }
  euclidean_norm_pcamicrobiome_100<-c(euclidean_norm_pcamicrobiome_100,sqrt(sum_distance))
  print(paste("The total distance",i,"is",sqrt(sum_distance),"For",(n-1),"axis"))
  i=i+1
}
print(euclidean_norm_pcamicrobiome_100)
newdf_pcamicrobiome_100$euclidean_norm_pcamicrobiome_100<-euclidean_norm_pcamicrobiome_100 #View(newdf_pcamicrobiome_100)

#col cumdistance
cumdistance_pcamicrobiome_100<-as.numeric(0)
i=2
while (i<=nrow(newdf_pcamicrobiome_100)) {
  distancep_100<-sum(((newdf_pcamicrobiome_100$euclidean_norm_pcamicrobiome_100[i]/sum(euclidean_norm_pcamicrobiome_100))*100)+cumdistance_pcamicrobiome_100[i-1])
  cumdistance_pcamicrobiome_100<-c(cumdistance_pcamicrobiome_100,distancep_100)
  i=i+1
}
print(cumdistance_pcamicrobiome_100)
newdf_pcamicrobiome_100$cumdistance_pcamicrobiome_100<-cumdistance_pcamicrobiome_100 #View(newdf_pcamicrobiome_100)

```
##|
## METABOLOME

The same process will be applied to the metabolome data. 
The code is only adapted for mixOmics-friendly data.

### PCA

```{r}
PCAGut_logmedaka.pca<-pca(G_log) #View(PCAGut_logmedaka.pca$variates$X)
```

### AXIS NEEDED

Let's determined the number of Axis needed downstream. 

```{r}
ExplVarTot_metabo=as.data.frame(PCAGut_logmedaka.pca$cum.var)[nrow(as.data.frame(PCAGut_logmedaka.pca$cum.var)),1] #View(ExplVarTot_metabo)

if (ExplVarTot_metabo<Threshold) {
  ExplVarTot_metabo=0
  nb_axis_metabo=1
  while (ExplVarTot_metabo<Threshold) {
    nb_axis_metabo=nb_axis_metabo+1
    PCAGut_logmedaka.pca<-pca(G_log,ncomp=nb_axis_metabo)
    ExplVarTot_metabo=as.data.frame(PCAGut_logmedaka.pca$cum.var)[nrow(as.data.frame(PCAGut_logmedaka.pca$cum.var)),1]
    Axis_needed_metabo<-nb_axis_metabo
    Axis_needed_end_metabo<-nb_axis_metabo+2
  }
  print(paste(nb_axis_metabo,"Axis are needed to explain",ExplVarTot_metabo,"of the total variance"))
  ExplVarTot_metabo=0
  nb_axis_metabo=0
}
```

### CENTROIDS CORDINATES

Make a new data frame containing all coordinates and the metadata needed and
split the dataset according to condition.
A reduced dataset can be created by retaining only the necessary axis fixed earlier.

```{r}
newdf_metabo<-data.frame(MetadataGUT_log$Cond2,MetadataGUT_log$Treatment,PCAGut_logmedaka.pca$variates$X) #View(newdf_metabo)
newdf_metabo_split<-split(newdf_metabo,f = MetadataGUT_log$Treatment) 
#View(newdf_metabo_split$`0`) 
#View(newdf_metabo_split$`100`)

newdf_metabo_0<-newdf_metabo_split$`0` %>% group_by(newdf_metabo_split$`0`$MetadataGUT_log.Cond2) %>%
  dplyr::summarise(across(starts_with("PC"),.fns=distance_metrics)) #View(newdf_metabo_0)

newdf_metabo_100<-newdf_metabo_split$`100` %>% group_by(newdf_metabo_split$`100`$MetadataGUT_log.Cond2) %>%
  dplyr::summarise(across(starts_with("PC"),.fns=distance_metrics)) #View(newdf_metabo_100)

#ajouter le d0_0 au dataframe exposé
names(newdf_metabo_0)[1] <-"Time_Exposure"
names(newdf_metabo_100)[1] <-"Time_Exposure"
newdf_metabo_100<-rbind(newdf_metabo_0[1,],newdf_metabo_100) #View(newdf_metabo_100)
```

### EUCLIDEAN NORM

The distances can then be processed by computing the euclidean norm between each group for all treatemnts.
Let's start with the non-exposed/control samples.

#### df_metabo_0

Three new columns will be added : 
  - a name/label for all cumputed-distances
  - the euclidean norm
  - the cumulative euclidean norm in %

```{r}
#col distance_label
Nlabel_metabo=1
distance_label_metabo_0<-as.character(newdf_metabo_0[1,1]) #View(distance_label_metabo_0)
while (Nlabel_metabo<nrow(newdf_metabo_0)) {
  distance_label2_metabo_0<-as.character(paste(newdf_metabo_0[Nlabel_metabo,1],"_",newdf_metabo_0[Nlabel_metabo+1,1]))
  distance_label_metabo_0<-c(distance_label_metabo_0,distance_label2_metabo_0)
  Nlabel_metabo=Nlabel_metabo+1
}
newdf_metabo_0$distance_label<-distance_label_metabo_0 #View(newdf_metabo_0)

#col euclidean_norm
euclidean_norm_metabo_0<-as.numeric(0) #View(euclidean_norm_metabo_0)
i=2
while (i<=nrow(newdf_metabo_0)) {
  sum_distance=0
  for (n in 2:(Axis_needed_metabo+1)) {
    distance_0<-sum((newdf_metabo_0[i,n]-newdf_metabo_0[i-1,n])^2)
    sum_distance=sum_distance+distance_0
  }
  euclidean_norm_metabo_0<-c(euclidean_norm_metabo_0,sqrt(sum_distance))
  print(paste("The total distance",i,"is",sqrt(sum_distance),"For",(n-1),"axis"))
  i=i+1
}
print(euclidean_norm_metabo_0)
newdf_metabo_0$euclidean_norm_metabolome_0<-euclidean_norm_metabo_0 #View(newdf_metabo_0)

#col cumdistance
cumdistance_metabo_0<-as.numeric(0)
i=2
while (i<=nrow(newdf_metabo_0)) {
  distancep_0<-sum(((newdf_metabo_0$euclidean_norm_metabolome_0[i]/sum(euclidean_norm_metabo_0))*100)+cumdistance_metabo_0[i-1])
  cumdistance_metabo_0<-c(cumdistance_metabo_0,distancep_0)
  i=i+1
}
print(cumdistance_metabo_0)
newdf_metabo_0$cumdistance_metabolome_0<-cumdistance_metabo_0 #View(newdf_metabo_0)
```

#### df_metabo_100

Let's perform the same process  to the exposed-samples.

```{r}
#col distance_label
Nlabel_metabo=1
distance_label_metabo_100<-as.character(newdf_metabo_100[1,1]) #View(distance_label_metabo_100)
while (Nlabel_metabo<nrow(newdf_metabo_100)) {
  distance_label2_metabo_100<-as.character(paste(newdf_metabo_100[Nlabel_metabo,1],"_",newdf_metabo_100[Nlabel_metabo+1,1]))
  distance_label_metabo_100<-c(distance_label_metabo_100,distance_label2_metabo_100)
  Nlabel_metabo=Nlabel_metabo+1
}
newdf_metabo_100$distance_label<-distance_label_metabo_100 #View(newdf_metabo_100)

#col euclidean_norm
euclidean_norm_metabo_100<-as.numeric(0) #View(euclidean_norm_metabo_100)
i=2
while (i<=nrow(newdf_metabo_100)) {
  sum_distance=0
  for (n in 2:(Axis_needed_metabo+1)) {
    distance_100<-sum((newdf_metabo_100[i,n]-newdf_metabo_100[i-1,n])^2)
    sum_distance=sum_distance+distance_100
  }
  euclidean_norm_metabo_100<-c(euclidean_norm_metabo_100,sqrt(sum_distance))
  print(paste("The total distance",i,"is",sqrt(sum_distance),"For",(n-1),"axis"))
  i=i+1
}
print(euclidean_norm_metabo_100)
newdf_metabo_100$euclidean_norm_metabolome_100<-euclidean_norm_metabo_100 #View(newdf_metabo_100)

#col cumdistance
cumdistance_metabo_100<-as.numeric(0)
i=2
while (i<=nrow(newdf_metabo_100)) {
  distancep_100<-sum(((newdf_metabo_100$euclidean_norm_metabolome_100[i]/sum(euclidean_norm_metabo_100))*100)+cumdistance_metabo_100[i-1])
  cumdistance_metabo_100<-c(cumdistance_metabo_100,distancep_100)
  i=i+1
}
print(cumdistance_metabo_100)
newdf_metabo_100$cumdistance_metabolome_100<-cumdistance_metabo_100 #View(newdf_metabo_100)
```
##|
## GRAPHICAL OUTPUT

#### add linear + caption

All necessary data are now acquired.
All the cumulative distances will be stored in a new dataframe and a few new columns
will be added for the graphical output.

```{r}
#col Time_label
time_label<-str_sub(newdf_pcamicrobiome_0$Time_Exposure,1,nchar(newdf_pcamicrobiome_0$Time_Exposure)-2)

newdf_cumdistance_all<-tibble(time_label,newdf_pcamicrobiome_0$cumdistance_pcamicrobiome_0,
                              newdf_pcamicrobiome_100$cumdistance_pcamicrobiome_100,
                              newdf_metabo_0$cumdistance_metabolome_0,
                              newdf_metabo_100$cumdistance_metabolome_100)

#col cumdistance_linear
time_point<-str_sub(newdf_pcamicrobiome_0$Time_Exposure,2,nchar(newdf_pcamicrobiome_0$Time_Exposure)-1)
time_point<-(str_sub(time_point,1,nchar(time_point)-1))
time_point<-as.numeric(time_point) #View(newdf_cumdistance_all)

#cumdistance_linear=NULL
nb_timepoint=nrow(newdf_cumdistance_all)
cumdistance_linear<-as.numeric(lapply(time_point,FUN=function(x) x*(100/nb_timepoint))) #View(cumdistance_linear)
newdf_cumdistance_all$cumdistance_linear<-cumdistance_linear 
#View(newdf_cumdistance_all)

distance_metrics_l_2PCA=as.list(distance_metrics)
parameters_2PCA<-paste(c("Microbiota = PCA-Eucl. ; Metabolome = PCA-Eucl. ;",
                    "\n\ndistance_metrics = ",
                    distance_metrics %>% as.list() %>%
                      .[[length(.)]] %>% .[[2]],
        " ; % of Expl. Var. = ",Threshold),collapse = "")
```

Here is the last final step : the ggplot code !

#### ggplot + pdf

```{r}
pdf(file=paste("Fig_Trajectory_","PCA-Eucl&PCA-Eucl_",
               distance_metrics %>% as.list() %>%
                 .[[length(.)]] %>% .[[2]]
               ,"_",Threshold,".pdf",sep="",collapse=NULL),
    width = 8.5, # The width of the plot in inches
    height = 6) # The height of the plot in inches

Fig_Trajectory_2PCA<-ggplot(newdf_cumdistance_all,
  aes(x=cumdistance_linear,y=cumdistance_linear))+
  geom_line(color="grey",size=1.5,linetype="dashed")+
  geom_line(aes(x=newdf_pcamicrobiome_0$cumdistance_pcamicrobiome_0,
                y=newdf_metabo_0$cumdistance_metabolome_0),color="blue",size=1)+
  geom_point(aes(x=newdf_pcamicrobiome_0$cumdistance_pcamicrobiome_0,
                 y=newdf_metabo_0$cumdistance_metabolome_0),
             shape=21,color="black",fill="#CBD5E8",size=3)+
  geom_label(aes(x=newdf_pcamicrobiome_0$cumdistance_pcamicrobiome_0,
                 y=newdf_metabo_0$cumdistance_metabolome_0,label=time_label),
             nudge_y=3,nudge_x=-4,size=5,color="white",fill="#80B1D3")+
  geom_line(aes(x=newdf_pcamicrobiome_100$cumdistance_pcamicrobiome_100,
                y=newdf_metabo_100$cumdistance_metabolome_100),color="#4DAF4A",size=1)+
  geom_point(aes(x=newdf_pcamicrobiome_100$cumdistance_pcamicrobiome_100,
                 y=newdf_metabo_100$cumdistance_metabolome_100),
             shape=21,color="black",fill="#CBD5E8",size=3.5)+
  geom_label(aes(x=newdf_pcamicrobiome_100$cumdistance_pcamicrobiome_100,
                 y=newdf_metabo_100$cumdistance_metabolome_100,label=time_label),
                nudge_y=3,nudge_x=-4,size=5,color="white",fill="darkgreen")+
  theme_bw()+expand_limits(y=c(0,100))+theme(aspect.ratio=1)+
  theme(plot.title=element_text(hjust = 0.5),plot.caption = element_text(hjust=0.5,size = 13),
        axis.text=element_text(size=16),axis.title=element_text(size=18))+
  labs(x="\nGut microbiota Kinetics\n",y="Gut metabolome Kinetics \n",caption=parameters_2PCA)

print(Fig_Trajectory_2PCA)
dev.off()
```

Thanks for reading !

##_________________________________________________________________________________________________